A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-10 paper.tex

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in the how important ARG-based methods become
across the wide range of predicted application
areas [CITE commentary pieces, tsinfer, ARGneedle any others?]
is how well inference methods and downstream analysis applications
interoperate. [Genetics doesn't have a good record here, let's
try not to make the same mistakes again.]

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity, but see below for confusion around this point)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\cite{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.

The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank}.

In retrospect, the choice of terminology around the succinct
tree sequence data structure was unfortunate, and has led to
significant confusion. [Discuss, and finish on a positive note]

>>>>

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package
% FIXME-  more explicit?
(e.g.\ Relate, ARGneedle)
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

The gARG encoding discussed and concretely defined in this manuscript
leads to highly efficient storage and processing of ARG data,
and has already been in use for several years.
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity, but see below for confusion around this point)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\citep{kelleher2016efficient} and subsequently been
extended and applied to forward-time
simulations~\citep{kelleher2018efficient,haller2018tree},
calculation of population genetics statistics~\citep{ralph2020efficiently}.
The succinct tree sequence encoding extends the basic definition
of a gARG provided here by stipulating a
simple tabular representation and nodes and edges,
and also defining a concise and lossless representation of
sequence variation using the ``site'' and  ``mutation'' tables.
The \texttt{tskit} library~\citep{tskit2023tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with gARGs (encoded as a succinct tree sequence).
Based on core functionality written
in C, it provides interfaces in C, Python and Rust.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.,][]{haller2019slim,speidel2019method,
adrion2020community,
terasaki2021geonomics,
baumdicker2021efficient,
fan2022genealogical,korfmann2022weak,
mahmoudi2022bayesian,petr2022slendr,rasmussen2022espalier,
zhang2023biobank}.

In retrospect, the choice of terminology around the succinct
tree sequence data structure was unfortunate, and has led to
significant confusion. As first introduced by~\cite{kelleher2016efficient}
a ``tree sequence`` was defined as the set of ``coalescence records``
that is output by Hudson's simulation algorithm.
This was then generalised to use the
tabular encoding described above and formally described as a ``succinct
tree sequence''~\citep{kelleher2018efficient}.
The ``succinct'' prefix here is intended to connect
with the concept of succinct data structures
\citep[e.g.][]{gog2014theory},
which have near-optimal
space usage but still support efficient retrieval and computation.
The concrete data structures of the succinct tree sequence, precisely
defined and with a particular focus on computational efficiency
did not have an obvious connection to the ARG definitions
provided by Griffiths and
colleagues~\citep{griffiths1991two,ethier1990two,
griffiths1996ancestral,griffiths1997ancestral}, and with the lack
of any other precise definitions, it seemed best to avoid the
term to prevent confusion. Unfortunately, however, the opposite
has occured. In describing the output of \tsinfer\ as a ``tree
sequence''~\citep{kelleher2019inferring} many
have---not unreasonably, but incorrectly---concluded that the
output is a collection of \emph{independent}
% MORE?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
trees~\citep[e.g.,][]{hejase2020summary,ignatieva2021kwarg}.
% Could say more here, but maybe we'll have a note about terminology
% in the discussion
